---
title: "Noah Jeffrey Analysis"
format: html
editor: visual
embed-resources: true
toc: true
code-fold: true
---

## Setup

```{r}
#| warning: false
#| message: false
library(tidyverse)
library(here)
library(lubridate)
library(lme4)
library(nlme)
library(car)
library(emmeans)
```

### Read in Data

```{r}
#| message: false
instructed <- read_csv(here("data", "instructed.csv"))
```

## Aggregating Data by Second

```{r}
instructed_agg <- instructed %>%
  mutate(sec = floor(time)) %>%
  group_by(participant, device_location, posture, sec) %>%
  summarize(angle_x = mean(angle_x), .groups = "keep")
```

```{r}
instructed_agg_nested <- instructed %>%
  mutate(nested_sec = floor(nested_time)) %>%
  group_by(participant, device_location, posture, nested_sec) %>%
  summarize(angle_x = mean(angle_x), .groups = "keep")
```

## Data Visualizations

```{r}
xmin <- c(0,10,14,24,27,37,41,51,54,64,67,77,80,90,93,103,106)
xmax <- lead(xmin,1)
xmax[17] <- 116
ymin <- rep(-70,17)
ymax <- rep(70,17)



time_ranges <- data.frame(
  xmin = xmin, 
  xmax = xmax, 
  ymin = ymin, 
  ymax = ymax,
  posture = c(rep(c("good", "transition", "bad", "transition"), 4), "good")) 
```

### Unaggregated data

```{r}
instructed %>%
  ggplot() +
  geom_line(aes(x = time, y = angle_x, color = device_location)) +
  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = posture), alpha = 0.4, data = time_ranges) +
  theme_minimal() +
  facet_wrap(~participant)

```

### Data aggregated by second

```{r}
instructed_agg %>%
  ggplot() +
  geom_line(aes(x = sec, y = angle_x, color = device_location)) +
  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = posture), alpha = 0.4, data = time_ranges) +
  theme_minimal() +
  facet_wrap(~participant)

```

### Single Participant Plot (aggregated by second)

```{r}
instructed_agg %>%
  filter(participant == 1) %>%
  ggplot() +
  geom_line(aes(x = sec, y = angle_x, color = device_location)) +
  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = posture), alpha = 0.4, data = time_ranges) +
  theme_minimal()
```

## Create Dataset for Model

```{r}
#| message: false
model_data <- instructed_agg_nested %>% 
  filter(posture != "transition") %>%
  group_by(participant, device_location, posture) %>%
  summarize(angle_x = mean(angle_x, na.rm = TRUE)) %>%
  ungroup() %>%
  pivot_wider(id_cols = c(participant, posture), names_from = device_location, values_from = angle_x) %>%
  mutate(posture = if_else(str_detect(posture, "good"), "good", posture)) %>%
  ungroup() %>%
  rename(
    "lower_back" = 3,
    "mid_back" = 4,
    "neck" = 5
  )
```

```{r}
model_data %>%
  ggplot() +
  geom_point(aes(x = posture, y = lower_back, color = factor(participant))) +
  facet_wrap(~participant) +
  theme(legend.position = "none",
        plot.title.position = "plot") +
  labs(title = "Lower Back Sensor") +
  theme_minimal()

model_data %>%
  ggplot() +
  geom_point(aes(x = posture, y = mid_back, color = factor(participant))) +
  facet_wrap(~participant) +
  theme(legend.position = "none",
        plot.title.position = "plot") +
  labs(title = "Mid Back Sensor") +
  theme_minimal()

model_data %>%
  ggplot() +
  geom_point(aes(x = posture, y = neck, color = factor(participant))) +
  facet_wrap(~participant) +
  theme(legend.position = "none",
        plot.title.position = "plot") +
  labs(title = "Neck Sensor") +
  theme_minimal()
```

## Model Testing

### Lower Back Sensor

#### Regular Mixed Model

```{r}
model_cd <- lmer(lower_back ~  posture + (1 + posture | participant), data = model_data)
S(model_cd)
Anova(model_cd)
plot(model_cd)

qqnorm(residuals(model_cd))
qqline(residuals(model_cd))

```

#### Generalized Linear Mixed Model

```{r}
glmer_cd <- glmer((lower_back + 6) ~  posture + (1 | participant), family = Gamma(link = "identity"), data = model_data)
S(glmer_cd)
plot(glmer_cd)
```

```{r}
#| code-fold: true
custom_combos <- data.frame(
  posture = model_data$posture %>% unique()
)

custom_preds <- ref_grid(
  model_cd,
  data = custom_combos,
  cov.reduce = FALSE,
)

model_preds <- emmeans(custom_preds, 
                         specs = ~ posture,
                         type = "response"
                         )


model_preds %>% 
  as_tibble() %>%
  ggplot(aes(color = posture)) +
  geom_point(aes(y = posture, x = emmean)) +
  geom_errorbar(aes(y = posture, xmin = lower.CL, xmax = upper.CL)) +
  geom_vline(aes(xintercept = xint), lty = 2, color = "red", data = data.frame(xint = 0)) +
  theme_minimal() +
  labs(title = "LMM: Lower Back Sensor") +
  theme(legend.position = "none",
        plot.title.position = "plot")
```

```{r}
#| code-fold: true
custom_combos <- data.frame(
  posture = model_data$posture %>% unique()
)

custom_preds <- ref_grid(
  glmer_cd,
  data = custom_combos,
  cov.reduce = FALSE,
)

model_preds <- emmeans(custom_preds, 
                         specs = ~ posture,
                         type = "response"
                         )


model_preds %>% 
  as_tibble() %>%
  ggplot(aes(color = posture)) +
  geom_point(aes(y = posture, x = response)) +
  geom_errorbar(aes(y = posture, xmin = asymp.LCL, xmax = asymp.UCL)) +
  geom_vline(aes(xintercept = xint), lty = 2, color = "red", data = data.frame(xint = 0)) +
  theme_minimal() +
  labs(title = "GLMM: Lower Back Sensor") +
  theme(legend.position = "none",
        plot.title.position = "plot")
```

### Mid Back Sensor

```{r}
model_data %>%
  select(participant, posture, mid_back) %>%
  filter(!is.na(mid_back)) %>%
  select(participant, posture) %>%
  distinct()
```

#### Regular Mixed Model

```{r}
model_ea <- lmer(mid_back ~ posture + (1  | participant), data = model_data)
S(model_ea)
Anova(model_ea)
plot(model_ea)


qqnorm(residuals(model_ea))
qqline(residuals(model_ea))
```

```{r}
model_ea <- lmer(mid_back ~ posture + (1  | participant), data = model_data)
S(model_ea)
Anova(model_ea)
plot(model_ea)


qqnorm(residuals(model_ea))
qqline(residuals(model_ea))
```

#### Generalized Linear Mixed Model

```{r}
glmer_ea <- glmer((-1*mid_back)+40 ~ posture + (1  | participant), family = Gamma(link = "identity"), data = model_data)
S(glmer_ea)
Anova(glmer_ea)
plot(glmer_ea)


glmer_ea <- glmer(mid_back+40 ~ posture + (1  | participant), family = Gamma(link = "identity"), data = model_data)
S(glmer_ea)
Anova(glmer_ea)
plot(glmer_ea)
```

```{r}
#| code-fold: true
custom_preds <- ref_grid(
  model_ea,
  data = custom_combos,
  cov.reduce = FALSE,
)

model_preds <- emmeans(custom_preds, 
                         specs = ~ posture,
                         type = "response"
                         )


model_preds %>% 
  as_tibble() %>%
  ggplot(aes(color = posture)) +
  geom_point(aes(y = posture, x = emmean)) +
  geom_errorbar(aes(y = posture, xmin = lower.CL, xmax = upper.CL)) +
  geom_vline(aes(xintercept = xint), lty = 2, color = "red", data = data.frame(xint = 0)) +
  theme_minimal() +
  labs(title = "LMM: Mid Back Sensor") +
  theme(legend.position = "none",
        plot.title.position = "plot")
```

```{r}
#| code-fold: true
custom_preds <- ref_grid(
  glmer_ea,
  data = custom_combos,
  cov.reduce = FALSE,
)

model_preds <- emmeans(custom_preds, 
                         specs = ~ posture,
                         type = "response"
                         )


model_preds %>% 
  as_tibble() %>%
  mutate(emmean = -1*(emmean-40),
         asymp.LCL = -1*(asymp.LCL-40),
         asymp.UCL = -1*(asymp.UCL-40)) %>%
  ggplot(aes(color = posture)) +
  geom_point(aes(y = posture, x = emmean)) +
  geom_errorbar(aes(y = posture, xmin = asymp.LCL, xmax = asymp.UCL)) +
  geom_vline(aes(xintercept = xint), lty = 2, color = "red", data = data.frame(xint = 0)) +
  theme_minimal() +
  labs(title = "GLMM: Mid Back Sensor") +
  theme(legend.position = "none",
        plot.title.position = "plot")
```

### Neck Sensor

#### Regular Mixed Model

```{r}

model_fe <- lmer(neck ~ posture + (1 + posture | participant), data = model_data)
S(model_fe)
Anova(model_fe)
plot(model_fe)


qqnorm(residuals(model_fe))
qqline(residuals(model_fe))
```

#### Generalized Linear Mixed Model

```{r}
glmer_fe <- glmer((neck*-1)+15 ~ posture + (1 | participant), family = Gamma(link = "identity"), data = model_data)
S(glmer_fe)
Anova(glmer_fe)
plot(glmer_fe)
```

```{r}
custom_preds <- ref_grid(
  model_fe,
  data = custom_combos,
  cov.reduce = FALSE,
)

model_preds <- emmeans(custom_preds, 
                         specs = ~ posture,
                         type = "response"
                         )


model_preds %>% 
  as_tibble() %>%
  ggplot(aes(color = posture)) +
  geom_point(aes(y = posture, x = emmean)) +
  geom_errorbar(aes(y = posture, xmin = lower.CL, xmax = upper.CL)) +
  geom_vline(aes(xintercept = xint), lty = 2, color = "red", data = data.frame(xint = 0)) +
  theme_minimal() +
  labs(title = "LMM: Neck Sensor") +
  theme(legend.position = "none",
        plot.title.position = "plot")
```

```{r}
custom_preds <- ref_grid(
  glmer_fe,
  data = custom_combos,
  cov.reduce = FALSE,
)

model_preds <- emmeans(custom_preds, 
                         specs = ~ posture,
                         type = "response"
                         )


model_preds %>% 
  as_tibble() %>%
  mutate(response = -1*(response-15),
       asymp.LCL = -1*(asymp.LCL-15),
       asymp.UCL = -1*(asymp.UCL-15)) %>%
  ggplot(aes(color = posture)) +
  geom_point(aes(y = posture, x = response)) +
  geom_errorbar(aes(y = posture, xmin = asymp.LCL, xmax = asymp.UCL)) +
  geom_vline(aes(xintercept = xint), lty = 2, color = "red", data = data.frame(xint = 0)) +
  theme_minimal() +
  labs(title = "GLMM: Neck Sensor") +
  theme(legend.position = "none",
        plot.title.position = "plot")
```

```{r}
model_data %>%
  pivot_longer(cols = lower_back:neck, names_to = "device", values_to = "angle_x") %>%
  ggplot() +
  geom_boxplot(aes(y = posture, x = angle_x, color = posture)) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title.position = "plot"
  ) +
  geom_vline(aes(xintercept = 0), color = "red", lty = 2) +
  facet_wrap(~device)

model_data %>%
  ggplot() +
  geom_boxplot(aes(y = posture, x = lower_back, color = posture)) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title.position = "plot"
  ) +
  geom_vline(aes(xintercept = 0), color = "red", lty = 2) +
  labs(title = "Device = Lower Back", x = "angle_x")

model_data %>%
  ggplot() +
  geom_boxplot(aes(y = posture, x = mid_back, color = posture)) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title.position = "plot"
  ) +
  geom_vline(aes(xintercept = 0), color = "red", lty = 2) +
  labs(title = "Device = Mid Back", x = "angle_x")


model_data %>%
  ggplot() +
  geom_boxplot(aes(y = posture, x = neck, color = posture)) +
  geom_point(aes(y = posture, x = neck, color = posture, shape = posture)) + 
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title.position = "plot"
  ) +
  geom_vline(aes(xintercept = 0), color = "red", lty = 2) +
  labs(title = "Device = Neck", x = "angle_x")
  
```

```{r}
library(ggridges)

model_data %>%
  pivot_longer(cols = lower_back:neck, names_to = "device", values_to = "angle_x") %>%
  ggplot() +
  ggridges::geom_density_line(aes(y = posture, x = angle_x, color = posture)) +
  # geom_boxplot(aes(y = posture, x = angle_x, color = posture)) +
  theme_minimal() +
  theme(
    legend.position = "none"
  ) +
  facet_wrap(~device) +
  geom_vline(aes(xintercept = 0), color = "red", lty = 2)
```
