---
title: "Spinal Segment Analysis"
author: "Brendan Callender and Andrew Kerr"
format: html
editor: source
embed-resources: true
code-fold: true
toc: true
---

*****************************************************************************************************************************
Use PI rather than CI 
Robinson not convinced with Gamma models
Classify Good, Meh, and Bad posture and highlight horizontally
*****************************************************************************************************************************

## Set-up

```{r}
#| warning: false
#| message: false
library(tidyverse)
library(here)
library(lubridate)
library(lme4)
library(nlme)
library(car)
library(emmeans)
library(ggridges)
library(viridis)
library(hrbrthemes)
```

## Data Read-in

```{r}
#| message: false
free <- read_csv(here("data", "free_15min.csv")) %>%
  mutate(
    participant = factor(
      participant, 
      levels = as.character(sort(as.numeric(unique(participant))))
      )
  ) %>%
  select(-angle_y, -angle_z)
instructed <- read_csv(here("data", "instructed.csv"))

free_norm_classified <- read_csv(here("data", "free_15min_classified.csv"))
time_ranges <- read_csv(here("data", "time_ranges_for_plotting.csv"))
```

### Data Cleaning

#### Min-max Normalization of Time for 15 Minutes Section

```{r}
free_norm <- free %>%
  group_by(participant, device_location) %>%
  mutate(
    # Calculate the time range for each group in seconds
    time_min_sec = as.numeric(min(chip_time)),
    time_max_sec = as.numeric(max(chip_time)),
    time_range_sec = time_max_sec - time_min_sec,
    # Normalize chip_time to be between 0 and 1
    normalized_chip_time = (as.numeric(chip_time) - time_min_sec) / time_range_sec
  ) %>%
  # If two or more normalized_chip_time values have multiple angle_x values, take the average
  group_by(participant, device_location, normalized_chip_time) %>%
  mutate(normalized_angle_x = mean(angle_x)) %>%
  select(-c(time_min_sec, time_max_sec, time_range_sec))
```

#### Aggregating Data by Second for Instructed Section

```{r}
instructed_agg <- instructed %>%
  mutate(sec = floor(time)) %>%
  group_by(participant, device_location, posture, sec) %>%
  summarize(angle_x = mean(angle_x), .groups = "keep")
```

```{r}
instructed_agg_nested <- instructed %>%
  mutate(nested_sec = floor(nested_time)) %>%
  group_by(participant, device_location, posture, nested_sec) %>%
  summarize(angle_x = mean(angle_x), .groups = "keep")
```

#### Create Dataset for Modeling

```{r}
#| message: false
model_data <- instructed_agg_nested %>% 
  filter(posture != "transition") %>%
  group_by(participant, device_location, posture) %>%
  summarize(angle_x = mean(angle_x, na.rm = TRUE)) %>%
  ungroup() %>%
  pivot_wider(id_cols = c(participant, posture), names_from = device_location, values_from = angle_x) %>%
  mutate(posture = if_else(str_detect(posture, "good"), "good", posture)) %>%
  ungroup() %>%
  rename(
    "lower_back" = 3,
    "mid_back" = 4,
    "neck" = 5
  )
```

## Data Visualization Settings

```{r}
xmin <- c(0,10,14,24,27,37,41,51,54,64,67,77,80,90,93,103,106)
xmax <- lead(xmin,1)
xmax[17] <- 116
ymin <- rep(-70,17)
ymax <- rep(70,17)

time_ranges <- data.frame(
  xmin = xmin, 
  xmax = xmax, 
  ymin = ymin, 
  ymax = ymax,
  posture = c(rep(c("good", "transition", "bad", "transition"), 4), "good")) 
```

## Summary Visualization for 15 Minutes Section

```{r}
summary_df <- free_norm %>%
  group_by(participant, device_location) %>%
  summarise(
    range_of_motion = max(angle_x) - min(angle_x),
    mean = mean(angle_x),
    median = median(angle_x),
    iqr = IQR(angle_x),
    .groups = 'keep'
  )

summary_df %>%
  ggplot() +
  geom_tile(aes(x = device_location, y = participant, fill = range_of_motion)) +
  scale_fill_distiller(palette = 'Spectral') +
  labs(title = "Heatmap of Range of Motion by Participant") +
  theme_bw()

summary_df %>%
  ggplot() +
  geom_tile(aes(x = device_location, y = participant, fill = mean)) +
  scale_fill_distiller(palette = 'Spectral') +
  labs(title = "Heatmap of Average Angle X by Participant") +
  theme_bw()

summary_df %>%
  ggplot() +
  geom_tile(aes(x = device_location, y = participant, fill = median)) +
  scale_fill_distiller(palette = 'Spectral') +
  labs(title = "Heatmap of Median Angle X by Participant") +
  theme_bw()

summary_df %>%
  ggplot() +
  geom_tile(aes(x = device_location, y = participant, fill = iqr)) +
  scale_fill_distiller(palette = 'Spectral') +
  labs(title = "Heatmap of IQR of Angle X by Participant") +
  theme_bw()
```

## Models for Instrcuted Section

### Lower Back

```{r}
glmer_cd <- glmer((lower_back + 6) ~  posture + (1 | participant), family = Gamma(link = "identity"), data = model_data)

glmer_cd <- lmer(lower_back ~  posture + (1 | participant), data = model_data)

S(glmer_cd)
plot(glmer_cd)
```

```{r}
custom_combos <- data.frame(
  posture = model_data$posture %>% unique()
)

custom_preds <- ref_grid(
  glmer_cd,
  data = custom_combos,
  cov.reduce = FALSE,
)

model_preds_lb <- emmeans(custom_preds, 
                         specs = ~ posture,
                         type = "response"
                         ) %>% 
  as_tibble()

model_preds_lb_pi <- predict(custom_preds, 
                         specs = ~ posture,
                         type = "response",
                         interval = "prediction"
                         ) %>% 
  as_tibble()

plot(custom_preds, PIs = TRUE)

model_preds_lb  %>%
  ggplot(aes(color = posture)) +
  
  # confidence interval
  geom_point(aes(y = posture, x = emmean), size = 3) +
  # geom_errorbar(aes(y = posture, xmin = asymp.LCL, xmax = asymp.UCL)) +
  geom_errorbar(aes(y = posture, xmin = lower.CL, xmax = upper.CL), width = 0, linewidth = 1) +
  
  # prediction interval
  geom_point(data = model_preds_lb_pi, 
             aes(y = posture, x = prediction)) +
  geom_errorbar(data = model_preds_lb_pi,
                aes(y = posture, xmin = lower.PL, xmax = upper.PL), width = 0, linewidth = 1, alpha = .2) +
  
  geom_vline(aes(xintercept = xint), lty = 2, color = "red", data = data.frame(xint = 0)) +
  
  # observed
  geom_point(data = model_data,
             aes(x = lower_back,
                 y = posture),
             alpha = 0.2) +
  theme_minimal() +
  labs(
    title = "GLMM: Lower Back Sensor",
    x = "Angle X",
    y = "Posture"
    ) +
  theme(legend.position = "none",
        plot.title.position = "plot")
```

### Mid Back

```{r}
glmer_ea <- glmer(mid_back+40 ~ posture + (1  | participant), family = Gamma(link = "identity"), data = model_data)
S(glmer_ea)
Anova(glmer_ea)
plot(glmer_ea)
```

```{r}
custom_preds <- ref_grid(
  glmer_ea,
  data = custom_combos,
  cov.reduce = FALSE,
)

model_preds_mb <- emmeans(custom_preds, 
                         specs = ~ posture,
                         type = "response"
                         ) %>% 
  as_tibble()

model_preds_mb_pi <- predict(custom_preds, 
                         specs = ~ posture,
                         type = "response",
                         interval = "prediction"
                         ) %>% 
  as_tibble()

model_preds_mb %>%
  mutate(emmean = -1*(emmean-40),
         asymp.LCL = -1*(asymp.LCL-40),
         asymp.UCL = -1*(asymp.UCL-40)) %>%
  ggplot(aes(color = posture)) +
  geom_point(aes(y = posture, x = emmean)) +
  geom_errorbar(aes(y = posture, xmin = asymp.LCL, xmax = asymp.UCL)) +
  geom_vline(aes(xintercept = xint), lty = 2, color = "red", data = data.frame(xint = 0)) +
  theme_minimal() +
  labs(title = "GLMM: Mid Back Sensor") +
  theme(legend.position = "none",
        plot.title.position = "plot")
```

### Neck

```{r}
glmer_fe <- glmer((neck*-1)+15 ~ posture + (1 | participant), family = Gamma(link = "identity"), data = model_data)
S(glmer_fe)
Anova(glmer_fe)
plot(glmer_fe)
```

```{r}
custom_preds <- ref_grid(
  glmer_fe,
  data = custom_combos,
  cov.reduce = FALSE,
)

model_preds_n <- emmeans(custom_preds, 
                         specs = ~ posture,
                         type = "response"
                         ) %>% 
  as_tibble()


model_preds_n_pi <- predict(custom_preds, 
                         specs = ~ posture,
                         type = "response",
                         interval = "prediction"
                         ) %>% 
  as_tibble()

model_preds_n  %>%
  mutate(response = -1*(response-15),
       asymp.LCL = -1*(asymp.LCL-15),
       asymp.UCL = -1*(asymp.UCL-15)) %>%
  ggplot(aes(color = posture)) +
  geom_point(aes(y = posture, x = response)) +
  geom_errorbar(aes(y = posture, xmin = asymp.LCL, xmax = asymp.UCL)) +
  geom_vline(aes(xintercept = xint), lty = 2, color = "red", data = data.frame(xint = 0)) +
  theme_minimal() +
  labs(title = "GLMM: Neck Sensor") +
  theme(legend.position = "none",
        plot.title.position = "plot")
```

## Visualizations for 15 Minute Section

### All Participants

```{r}
free_norm %>%
  ggplot() +
    geom_line(aes(x = normalized_chip_time, y = normalized_angle_x, color = participant)) +
    facet_wrap(vars(device_location), nrow = 3, ncol = 1) +
    theme(
    strip.text = element_text(size = 8),  # Adjust label size
    axis.text = element_text(size = 6),   # Adjust axis text size
    plot.margin = unit(c(1, 1, 1, 1), "mm") # Adjust plot margins
  ) +
  labs(title = "Normalized Angle X by Normalized Chip Time") +
  theme_bw() +
  ylim(-50, 50)
```


```{r}
#| eval: false
posture_cis <- model_preds_lb %>%
  left_join(model_preds_mb, by = "posture", suffix = c("_lower_back", "_mid_back")) %>%
  left_join(model_preds_n, by = "posture", suffix = c("_lower_back", "_neck")) %>%
  rename(
    response_mid_back = emmean,
    asymp.LCL_neck = asymp.LCL,
    asymp.UCL_neck = asymp.UCL
  ) %>%
  select(posture, starts_with("response"), starts_with("asymp")) %>%
  pivot_longer(
    cols = -posture,
    names_to = c(".value", "location"),
    names_pattern = "(asymp\\.[LU]CL)(_.+)$"
  ) %>%
  rename(
    lower = asymp.LCL,
    upper = asymp.UCL
  ) %>%
  filter(!is.na(lower)) %>%
  mutate(location = gsub("^_", "", location)) %>%
  arrange(factor(posture, levels = posture_priority))

# Return all matching postures as a list column
classify_postures <- function(angles, locations, rules) {
  # Create all possible combinations
  expand_grid(
    angle = angles,
    loc = locations,
    posture = unique(rules$posture)
  ) %>%
  left_join(rules, by = c("loc" = "location", "posture")) %>%
  mutate(
    matches = angle >= lower & angle <= upper & !is.na(angle)
  ) %>%
  group_by(angle, loc) %>%
  summarise(
    postures = if(any(matches)) {
      paste(sort(unique(posture[matches])), collapse = "+")
    } else {
      "unknown"
    },
    .groups = "drop"
  ) %>%
  right_join(
    tibble(angle = angles, loc = locations),
    by = c("angle", "loc")
  ) %>%
  mutate(postures = coalesce(postures, "unknown")) %>%
  pull(postures)
}

# Apply classification and prepare data for plotting
free_norm_classified <- free_norm %>%
  mutate(
    classified_posture = classify_postures(
      angles = angle_x,
      locations = device_location,
      rules = posture_cis
    )
  )

time_ranges <- free_norm_classified %>%
  group_by(participant, device_location) %>%
  mutate(
    posture_change = classified_posture != lag(classified_posture, default = first(classified_posture)),
    segment_id = cumsum(posture_change)
  ) %>%
  group_by(participant, device_location, segment_id, classified_posture) %>%
  summarise(
    xmin = min(normalized_chip_time),
    xmax = max(normalized_chip_time),
    ymin = min(normalized_angle_x),
    ymax = max(normalized_angle_x),
    .groups = "drop"
  )

# Save data
write_csv(free_norm_classified, "data/free_15min_classified.csv")
write_csv(time_ranges, "data/time_ranges_for_plotting.csv")
```

```{r}
#| messsage: false
free_norm_classified <- read_csv(here("data", "free_15min_classified.csv"))
time_ranges <- read_csv(here("data", "time_ranges_for_plotting.csv"))
```

```{r}
plot_classifications <- function(df, PI, loc, p) {
  # Filter data for this participant and location
  plot_data <- df %>%
    filter(participant == p, device_location == loc)
  
  # Extract good posture prediction interval limits
  good_lower <- PI %>% filter(posture == "good") %>% pull(lower.PL)
  good_upper <- PI %>% filter(posture == "good") %>% pull(upper.PL)

  # TEMP Definition of Meh Posture
  meh_low_lower <- good_lower - 3
  meh_low_upper <- good_lower
  
  meh_high_lower <- good_upper
  meh_high_upper <- good_upper + 3
  
  # --- Calculate Posture Proportions ---
  plot_data_classified <- plot_data %>%
    mutate(
      # Classify each data point
      posture_group = case_when(
        normalized_angle_x >= good_lower & normalized_angle_x <= good_upper ~ "Good",
        (normalized_angle_x >= meh_low_lower & normalized_angle_x < good_lower) |
        (normalized_angle_x > good_upper & normalized_angle_x <= meh_high_upper) ~ "Meh",
        TRUE ~ "Bad"
      ),
      posture_group = factor(posture_group, levels = c("Good", "Meh", "Bad"))
    )
  
  # Calculate proportions for each posture group
  proportions_tbl <- plot_data_classified %>%
    group_by(posture_group) %>%
    summarise(proportion = n() / nrow(plot_data_classified)) %>%
    complete(posture_group, fill = list(proportion = 0)) %>%
    arrange(posture_group)

  # Create legend labels with proportions
  legend_labels_with_props <- paste0(
    proportions_tbl$posture_group, 
    " (", 
    scales::percent(proportions_tbl$proportion, accuracy = 1),
    ")"
  )
  # --- End Calculate Posture Proportions ---
  
  # Define x-axis limits for the background rectangles (assuming normalized time 0-1)
  xmin_rect <- 0
  xmax_rect <- 1
  
  # Define y-axis limits for the different posture classification zones
  ymin_rect <- c(-Inf, meh_low_lower, good_lower, good_upper, meh_high_upper)
  ymax_rect <- c(meh_low_lower, good_lower, good_upper, meh_high_upper, Inf)
  
  # Create a dataframe for the background rectangles representing posture zones
  time_ranges <- data.frame(
    xmin = xmin_rect,  
    xmax = xmax_rect,  
    ymin = ymin_rect,  
    ymax = ymax_rect,
    posture_cat = factor(c("Bad", "Meh", "Good", "Meh", "Bad"), 
                         levels = c("Good", "Meh", "Bad"))
  )
  
  # Generate the plot
  classification_plot <- ggplot(plot_data) +
    # Line plot for the actual data
    geom_line(aes(x = normalized_chip_time, y = normalized_angle_x)) +
    # Rectangles for posture zones
    geom_rect(
      aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = posture_cat), 
      alpha = 0.4, 
      data = time_ranges
    ) +
    theme_minimal() +
    labs(
      title = paste("Posture Classification for Participant", p),
      subtitle = paste("Location:", loc, "- 15min Free Section"),
      x = "Normalized Chip Time",
      y = "Normalized Angle X",
      fill = "Posture Type"
    ) +
    scale_fill_manual(
      values = c("Good" = "seagreen3", "Meh" = "gold", "Bad" = "salmon"),
      breaks = c("Good", "Meh", "Bad"),
      labels = legend_labels_with_props
    )
  
  print(proportions_tbl)
  
  return(classification_plot)

}
```

#### Lower Back

```{r}
plot_classifications(free_norm, model_preds_lb_pi, "lower_back", 3)
```

#### Lower Back

```{r}
plot_classifications(free_norm, model_preds_mb_pi, 3, "mid_back")
```

#### Lower Back

```{r}
plot_classifications(free_norm, model_preds_n_pi, 3, "neck")
```

### Ridge Plots

#### All Devices

```{r}
#| message: false
#| warning: false
free_norm %>%
  ggplot(aes(x = angle_x, y = participant, fill = device_location)) +
    geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
    scale_fill_viridis(
      option = "D", 
      discrete = TRUE,
      labels = c(
        "lower_back" = "Lower Back",
        "mid_back" = "Mid Back",
        "neck" = "Neck"
        )
      ) +
    labs(
      title = 'Distribution of Time Spent in Normalized Angle X by Participant',
      fill = 'Device Type',
      x = 'Angle X',
      y = 'Participant'
      ) +
    theme_ipsum() +
      theme(
        panel.spacing = unit(0.1, "lines"),
        strip.text.x = element_text(size = 8)
      ) +
    xlim(-50, 70)
```

#### Lower Back

```{r}
#| message: false
#| warning: false
free_norm %>%
  filter(device_location == 'lower_back') %>%
  ggplot(aes(x = angle_x, y = participant, fill = ..x..)) +
    geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
    scale_fill_viridis(option = "C") +
    labs(
      title = 'Distribution of Time Spent in Normalized Angle X by Participant', 
      subtitle = 'Lower Back',
      x = 'Angle X',
      y = 'Participant'
      ) +
    theme_ipsum() +
      theme(
        legend.position="none",
        panel.spacing = unit(0.1, "lines"),
        strip.text.x = element_text(size = 8)
      ) +
    xlim(-50, 70)
```

#### Mid Back

```{r}
#| message: false
#| warning: false
free_norm %>%
  filter(device_location == 'mid_back') %>%
  ggplot(aes(x = angle_x, y = participant, fill = ..x..)) +
    geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
    scale_fill_viridis(option = "C") +
    labs(
      title = 'Distribution of Time Spent in Normalized Angle X by Participant', 
      subtitle = 'Mid Back',
      x = 'Angle X',
      y = 'Participant'
      ) +
    theme_ipsum() +
      theme(
        legend.position="none",
        panel.spacing = unit(0.1, "lines"),
        strip.text.x = element_text(size = 8)
      ) +
    xlim(-50, 70)
```

#### Neck

```{r}
#| message: false
#| warning: false
free_norm %>%
  filter(device_location == 'neck') %>%
  ggplot(aes(x = angle_x, y = participant, fill = ..x..)) +
    geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
    scale_fill_viridis(option = "C") +
    labs(
      title = 'Distribution of Time Spent in Normalized Angle X by Participant', 
      subtitle = 'Neck',
      x = 'Angle X',
      y = 'Participant'
      ) +
    theme_ipsum() +
      theme(
        legend.position="none",
        panel.spacing = unit(0.1, "lines"),
        strip.text.x = element_text(size = 8)
      ) +
    xlim(-50, 70)
```



