---
title: "Spinal Segment Analysis"
author: "Brendan Callender and Andrew Kerr"
format: html
editor: source
embed-resources: true
code-fold: true
toc: true
---

## Set-up

```{r}
#| warning: false
#| message: false
library(tidyverse)
library(here)
library(lubridate)
library(lme4)
library(nlme)
library(car)
library(emmeans)
library(ggridges)
library(viridis)
library(hrbrthemes)
```

## Data Read-in

```{r}
#| message: false
free <- read_csv(here("data", "free_15min.csv")) %>%
  mutate(
    participant = factor(
      participant, 
      levels = as.character(sort(as.numeric(unique(participant))))
      )
  ) %>%
  select(-angle_y, -angle_z)
instructed <- read_csv(here("data", "instructed.csv"))
```

### Data Cleaning

#### Min-max Normalization of Time for 15 Minutes Section

```{r}
free_norm <- free %>%
  group_by(participant, device_location) %>%
  mutate(
    # Calculate the time range for each group in seconds
    time_min_sec = as.numeric(min(chip_time)),
    time_max_sec = as.numeric(max(chip_time)),
    time_range_sec = time_max_sec - time_min_sec,
    # Normalize chip_time to be between 0 and 1
    normalized_chip_time = (as.numeric(chip_time) - time_min_sec) / time_range_sec
  ) %>%
  # If two or more normalized_chip_time values have multiple angle_x values, take the average
  group_by(participant, device_location, normalized_chip_time) %>%
  mutate(normalized_angle_x = mean(angle_x)) %>%
  select(-c(time_min_sec, time_max_sec, time_range_sec))
```

#### Aggregating Data by Second for Instructed Section

```{r}
instructed_agg <- instructed %>%
  mutate(sec = floor(time)) %>%
  group_by(participant, device_location, posture, sec) %>%
  summarize(angle_x = mean(angle_x), .groups = "keep")
```

```{r}
instructed_agg_nested <- instructed %>%
  mutate(nested_sec = floor(nested_time)) %>%
  group_by(participant, device_location, posture, nested_sec) %>%
  summarize(angle_x = mean(angle_x), .groups = "keep")
```

#### Create Dataset for Modeling

```{r}
#| message: false
model_data <- instructed_agg_nested %>% 
  filter(posture != "transition") %>%
  group_by(participant, device_location, posture) %>%
  summarize(angle_x = mean(angle_x, na.rm = TRUE)) %>%
  ungroup() %>%
  pivot_wider(id_cols = c(participant, posture), names_from = device_location, values_from = angle_x) %>%
  mutate(posture = if_else(str_detect(posture, "good"), "good", posture)) %>%
  ungroup() %>%
  rename(
    "lower_back" = 3,
    "mid_back" = 4,
    "neck" = 5
  )
```

## Data Visualization Settings

```{r}
xmin <- c(0,10,14,24,27,37,41,51,54,64,67,77,80,90,93,103,106)
xmax <- lead(xmin,1)
xmax[17] <- 116
ymin <- rep(-70,17)
ymax <- rep(70,17)

time_ranges <- data.frame(
  xmin = xmin, 
  xmax = xmax, 
  ymin = ymin, 
  ymax = ymax,
  posture = c(rep(c("good", "transition", "bad", "transition"), 4), "good")) 
```

## Summary Visualization for 15 Minutes Section

```{r}
summary_df <- free_norm %>%
  group_by(participant, device_location) %>%
  summarise(
    range_of_motion = max(angle_x) - min(angle_x),
    mean = mean(angle_x),
    median = median(angle_x),
    iqr = IQR(angle_x),
    .groups = 'keep'
  )

summary_df %>%
  ggplot() +
  geom_tile(aes(x = device_location, y = participant, fill = range_of_motion)) +
  scale_fill_distiller(palette = 'Spectral') +
  labs(title = "Heatmap of Range of Motion by Participant") +
  theme_bw()

summary_df %>%
  ggplot() +
  geom_tile(aes(x = device_location, y = participant, fill = mean)) +
  scale_fill_distiller(palette = 'Spectral') +
  labs(title = "Heatmap of Average Angle X by Participant") +
  theme_bw()

summary_df %>%
  ggplot() +
  geom_tile(aes(x = device_location, y = participant, fill = median)) +
  scale_fill_distiller(palette = 'Spectral') +
  labs(title = "Heatmap of Median Angle X by Participant") +
  theme_bw()

summary_df %>%
  ggplot() +
  geom_tile(aes(x = device_location, y = participant, fill = iqr)) +
  scale_fill_distiller(palette = 'Spectral') +
  labs(title = "Heatmap of IQR of Angle X by Participant") +
  theme_bw()
```

## Models for Instrcuted Section

### Lower Back Sensor

### Mid Back Sensor

### Neck Sensor

## Visualizations for 15 Minute Section

### All Participants

```{r}
free_norm %>%
  # filter(participant_number %in% seq(1, 5)) %>%
  ggplot() +
    geom_line(aes(x = normalized_chip_time, y = normalized_angle_x, color = participant)) +
    facet_wrap(vars(device_location), nrow = 3, ncol = 1) +
    theme(
    strip.text = element_text(size = 8),  # Adjust label size
    axis.text = element_text(size = 6),   # Adjust axis text size
    plot.margin = unit(c(1, 1, 1, 1), "mm") # Adjust plot margins
  ) +
  labs(title = "Normalized Angle X by Normalized Chip Time") +
  theme_bw()

free_norm %>%
  # filter(participant_number %in% seq(1, 5)) %>%
  ggplot() +
    geom_line(aes(x = normalized_chip_time, y = normalized_angle_x, color = participant)) +
    facet_wrap(vars(device_location), nrow = 3, ncol = 1) +
    theme(
    strip.text = element_text(size = 8),  # Adjust label size
    axis.text = element_text(size = 6),   # Adjust axis text size
    plot.margin = unit(c(1, 1, 1, 1), "mm") # Adjust plot margins
  ) +
  labs(title = "Normalized Angle X by Normalized Chip Time") +
  theme_bw() +
  ylim(-50, 50)
```

### Ridge Plots

```{r}
#| message: false
#| warning: false
free_norm %>%
  ggplot(aes(x = normalized_angle_x, y = participant, fill = device_location)) +
    geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
    scale_fill_viridis(name = "angle_x", option = "D", discrete = TRUE) +
    labs(title = 'Distribution of Time Spent in Normalized Angle X by Participant') +
    theme_ipsum() +
      theme(
        legend.position="none",
        panel.spacing = unit(0.1, "lines"),
        strip.text.x = element_text(size = 8)
      ) +
    xlim(-50, 70)

free_norm %>%
  filter(device_location == 'lower_back') %>%
  ggplot(aes(x = normalized_angle_x, y = participant, fill = ..x..)) +
    geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
    scale_fill_viridis(name = "angle_x", option = "C") +
    labs(title = 'Distribution of Time Spent in Normalized Angle X by Participant', subtitle = 'Lower Back') +
    theme_ipsum() +
      theme(
        legend.position="none",
        panel.spacing = unit(0.1, "lines"),
        strip.text.x = element_text(size = 8)
      ) +
    xlim(-50, 70)

free_norm %>%
  filter(device_location == 'mid_back') %>%
  ggplot(aes(x = normalized_angle_x, y = participant, fill = ..x..)) +
    geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
    scale_fill_viridis(name = "angle_x", option = "C") +
    labs(title = 'Distribution of Time Spent in Normalized Angle X by Participant', subtitle = 'Mid Back') +
    theme_ipsum() +
      theme(
        legend.position="none",
        panel.spacing = unit(0.1, "lines"),
        strip.text.x = element_text(size = 8)
      ) +
    xlim(-50, 70)

free_norm %>%
  filter(device_location == 'neck') %>%
  ggplot(aes(x = normalized_angle_x, y = participant, fill = ..x..)) +
    geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
    scale_fill_viridis(name = "angle_x", option = "C") +
    labs(title = 'Distribution of Time Spent in Normalized Angle X by Participant', subtitle = 'Neck') +
    theme_ipsum() +
      theme(
        legend.position="none",
        panel.spacing = unit(0.1, "lines"),
        strip.text.x = element_text(size = 8)
      ) +
    xlim(-50, 70)
```



