---
title: "Spinal Segment Analysis"
author: "Brendan Callender and Andrew Kerr"
format: html
editor: source
embed-resources: true
code-fold: true
toc: true
---

*****************************************************************************************************************************
NOTE: we are currently using Gamma, so only positive angle values, but we want to take into account negative as well
*****************************************************************************************************************************

## Set-up

```{r}
#| warning: false
#| message: false
library(tidyverse)
library(here)
library(lubridate)
library(lme4)
library(nlme)
library(car)
library(emmeans)
library(ggridges)
library(viridis)
library(hrbrthemes)
```

## Data Read-in

```{r}
#| message: false
free <- read_csv(here("data", "free_15min.csv")) %>%
  mutate(
    participant = factor(
      participant, 
      levels = as.character(sort(as.numeric(unique(participant))))
      )
  ) %>%
  select(-angle_y, -angle_z)
instructed <- read_csv(here("data", "instructed.csv"))

free_norm_classified <- read_csv(here("data", "free_15min_classified.csv"))
time_ranges <- read_csv(here("data", "time_ranges_for_plotting.csv"))
```

### Data Cleaning

#### Min-max Normalization of Time for 15 Minutes Section

```{r}
free_norm <- free %>%
  group_by(participant, device_location) %>%
  mutate(
    # Calculate the time range for each group in seconds
    time_min_sec = as.numeric(min(chip_time)),
    time_max_sec = as.numeric(max(chip_time)),
    time_range_sec = time_max_sec - time_min_sec,
    # Normalize chip_time to be between 0 and 1
    normalized_chip_time = (as.numeric(chip_time) - time_min_sec) / time_range_sec
  ) %>%
  # If two or more normalized_chip_time values have multiple angle_x values, take the average
  group_by(participant, device_location, normalized_chip_time) %>%
  mutate(normalized_angle_x = mean(angle_x)) %>%
  select(-c(time_min_sec, time_max_sec, time_range_sec))
```

#### Aggregating Data by Second for Instructed Section

```{r}
instructed_agg <- instructed %>%
  mutate(sec = floor(time)) %>%
  group_by(participant, device_location, posture, sec) %>%
  summarize(angle_x = mean(angle_x), .groups = "keep")
```

```{r}
instructed_agg_nested <- instructed %>%
  mutate(nested_sec = floor(nested_time)) %>%
  group_by(participant, device_location, posture, nested_sec) %>%
  summarize(angle_x = mean(angle_x), .groups = "keep")
```

#### Create Dataset for Modeling

```{r}
#| message: false
model_data <- instructed_agg_nested %>% 
  filter(posture != "transition") %>%
  group_by(participant, device_location, posture) %>%
  summarize(angle_x = mean(angle_x, na.rm = TRUE)) %>%
  ungroup() %>%
  pivot_wider(id_cols = c(participant, posture), names_from = device_location, values_from = angle_x) %>%
  mutate(posture = if_else(str_detect(posture, "good"), "good", posture)) %>%
  ungroup() %>%
  rename(
    "lower_back" = 3,
    "mid_back" = 4,
    "neck" = 5
  )
```

## Data Visualization Settings

```{r}
xmin <- c(0,10,14,24,27,37,41,51,54,64,67,77,80,90,93,103,106)
xmax <- lead(xmin,1)
xmax[17] <- 116
ymin <- rep(-70,17)
ymax <- rep(70,17)

time_ranges <- data.frame(
  xmin = xmin, 
  xmax = xmax, 
  ymin = ymin, 
  ymax = ymax,
  posture = c(rep(c("good", "transition", "bad", "transition"), 4), "good")) 
```

## Summary Visualization for 15 Minutes Section

```{r}
summary_df <- free_norm %>%
  group_by(participant, device_location) %>%
  summarise(
    range_of_motion = max(angle_x) - min(angle_x),
    mean = mean(angle_x),
    median = median(angle_x),
    iqr = IQR(angle_x),
    .groups = 'keep'
  )

summary_df %>%
  ggplot() +
  geom_tile(aes(x = device_location, y = participant, fill = range_of_motion)) +
  scale_fill_distiller(palette = 'Spectral') +
  labs(title = "Heatmap of Range of Motion by Participant") +
  theme_bw()

summary_df %>%
  ggplot() +
  geom_tile(aes(x = device_location, y = participant, fill = mean)) +
  scale_fill_distiller(palette = 'Spectral') +
  labs(title = "Heatmap of Average Angle X by Participant") +
  theme_bw()

summary_df %>%
  ggplot() +
  geom_tile(aes(x = device_location, y = participant, fill = median)) +
  scale_fill_distiller(palette = 'Spectral') +
  labs(title = "Heatmap of Median Angle X by Participant") +
  theme_bw()

summary_df %>%
  ggplot() +
  geom_tile(aes(x = device_location, y = participant, fill = iqr)) +
  scale_fill_distiller(palette = 'Spectral') +
  labs(title = "Heatmap of IQR of Angle X by Participant") +
  theme_bw()
```

## Models for Instrcuted Section

### Lower Back

```{r}
glmer_cd <- glmer((lower_back + 6) ~  posture + (1 | participant), family = Gamma(link = "identity"), data = model_data)
S(glmer_cd)
plot(glmer_cd)
```

```{r}
custom_combos <- data.frame(
  posture = model_data$posture %>% unique()
)

custom_preds <- ref_grid(
  glmer_cd,
  data = custom_combos,
  cov.reduce = FALSE,
)

model_preds_lb <- emmeans(custom_preds, 
                         specs = ~ posture,
                         type = "response"
                         ) %>% 
  as_tibble()

model_preds_lb  %>%
  ggplot(aes(color = posture)) +
  geom_point(aes(y = posture, x = response)) +
  geom_errorbar(aes(y = posture, xmin = asymp.LCL, xmax = asymp.UCL)) +
  geom_vline(aes(xintercept = xint), lty = 2, color = "red", data = data.frame(xint = 0)) +
  theme_minimal() +
  labs(
    title = "GLMM: Lower Back Sensor",
    x = "Angle X",
    y = "Posture"
    ) +
  theme(legend.position = "none",
        plot.title.position = "plot")
```

### Mid Back

```{r}
glmer_ea <- glmer(mid_back+40 ~ posture + (1  | participant), family = Gamma(link = "identity"), data = model_data)
S(glmer_ea)
Anova(glmer_ea)
plot(glmer_ea)
```

```{r}
custom_preds <- ref_grid(
  glmer_ea,
  data = custom_combos,
  cov.reduce = FALSE,
)

model_preds_mb <- emmeans(custom_preds, 
                         specs = ~ posture,
                         type = "response"
                         ) %>% 
  as_tibble()

model_preds_mb %>%
  mutate(emmean = -1*(emmean-40),
         asymp.LCL = -1*(asymp.LCL-40),
         asymp.UCL = -1*(asymp.UCL-40)) %>%
  ggplot(aes(color = posture)) +
  geom_point(aes(y = posture, x = emmean)) +
  geom_errorbar(aes(y = posture, xmin = asymp.LCL, xmax = asymp.UCL)) +
  geom_vline(aes(xintercept = xint), lty = 2, color = "red", data = data.frame(xint = 0)) +
  theme_minimal() +
  labs(title = "GLMM: Mid Back Sensor") +
  theme(legend.position = "none",
        plot.title.position = "plot")
```

### Neck

```{r}
glmer_fe <- glmer((neck*-1)+15 ~ posture + (1 | participant), family = Gamma(link = "identity"), data = model_data)
S(glmer_fe)
Anova(glmer_fe)
plot(glmer_fe)
```

```{r}
custom_preds <- ref_grid(
  glmer_fe,
  data = custom_combos,
  cov.reduce = FALSE,
)

model_preds_n <- emmeans(custom_preds, 
                         specs = ~ posture,
                         type = "response"
                         ) %>% 
  as_tibble()


model_preds_n  %>%
  mutate(response = -1*(response-15),
       asymp.LCL = -1*(asymp.LCL-15),
       asymp.UCL = -1*(asymp.UCL-15)) %>%
  ggplot(aes(color = posture)) +
  geom_point(aes(y = posture, x = response)) +
  geom_errorbar(aes(y = posture, xmin = asymp.LCL, xmax = asymp.UCL)) +
  geom_vline(aes(xintercept = xint), lty = 2, color = "red", data = data.frame(xint = 0)) +
  theme_minimal() +
  labs(title = "GLMM: Neck Sensor") +
  theme(legend.position = "none",
        plot.title.position = "plot")
```

## Visualizations for 15 Minute Section

### All Participants

```{r}
free_norm %>%
  ggplot() +
    geom_line(aes(x = normalized_chip_time, y = normalized_angle_x, color = participant)) +
    facet_wrap(vars(device_location), nrow = 3, ncol = 1) +
    theme(
    strip.text = element_text(size = 8),  # Adjust label size
    axis.text = element_text(size = 6),   # Adjust axis text size
    plot.margin = unit(c(1, 1, 1, 1), "mm") # Adjust plot margins
  ) +
  labs(title = "Normalized Angle X by Normalized Chip Time") +
  theme_bw() +
  ylim(-50, 50)
```


```{r}
#| eval: false
posture_cis <- model_preds_lb %>%
  left_join(model_preds_mb, by = "posture", suffix = c("_lower_back", "_mid_back")) %>%
  left_join(model_preds_n, by = "posture", suffix = c("_lower_back", "_neck")) %>%
  rename(
    response_mid_back = emmean,
    asymp.LCL_neck = asymp.LCL,
    asymp.UCL_neck = asymp.UCL
  ) %>%
  select(posture, starts_with("response"), starts_with("asymp")) %>%
  pivot_longer(
    cols = -posture,
    names_to = c(".value", "location"),
    names_pattern = "(asymp\\.[LU]CL)(_.+)$"
  ) %>%
  rename(
    lower = asymp.LCL,
    upper = asymp.UCL
  ) %>%
  filter(!is.na(lower)) %>%
  mutate(location = gsub("^_", "", location)) %>%
  arrange(factor(posture, levels = posture_priority))

# Return all matching postures as a list column
classify_postures <- function(angles, locations, rules) {
  # Create all possible combinations
  expand_grid(
    angle = angles,
    loc = locations,
    posture = unique(rules$posture)
  ) %>%
  left_join(rules, by = c("loc" = "location", "posture")) %>%
  mutate(
    matches = angle >= lower & angle <= upper & !is.na(angle)
  ) %>%
  group_by(angle, loc) %>%
  summarise(
    postures = if(any(matches)) {
      paste(sort(unique(posture[matches])), collapse = "+")
    } else {
      "unknown"
    },
    .groups = "drop"
  ) %>%
  right_join(
    tibble(angle = angles, loc = locations),
    by = c("angle", "loc")
  ) %>%
  mutate(postures = coalesce(postures, "unknown")) %>%
  pull(postures)
}

# Apply classification and prepare data for plotting
free_norm_classified <- free_norm %>%
  mutate(
    classified_posture = classify_postures(
      angles = angle_x,
      locations = device_location,
      rules = posture_cis
    )
  )

time_ranges <- free_norm_classified %>%
  group_by(participant, device_location) %>%
  mutate(
    posture_change = classified_posture != lag(classified_posture, default = first(classified_posture)),
    segment_id = cumsum(posture_change)
  ) %>%
  group_by(participant, device_location, segment_id, classified_posture) %>%
  summarise(
    xmin = min(normalized_chip_time),
    xmax = max(normalized_chip_time),
    ymin = min(normalized_angle_x),
    ymax = max(normalized_angle_x),
    .groups = "drop"
  )

# Save data
write_csv(free_norm_classified, "data/free_15min_classified.csv")
write_csv(time_ranges, "data/time_ranges_for_plotting.csv")
```

```{r}
plot_classifications <- function(free_norm_classified, time_ranges, p, loc) {
  # Filter data for this participant and location
  plot_data <- free_norm_classified %>%
    filter(participant == p, device_location == loc)
  
  # Get all unique posture combinations in the data
  all_postures <- time_ranges %>%
    filter(participant == p, device_location == loc) %>%
    pull(classified_posture) %>%
    unique()
  
  # Calculate proportions
  proportions <- time_ranges %>%
    filter(participant == p, device_location == loc) %>%
    mutate(duration = xmax - xmin) %>%
    group_by(classified_posture) %>%
    summarise(
      proportion = sum(duration) / sum(.$duration)
    )
  
  # Base colors for single postures
  base_colors <- c(
    "good" = "#009E73",
    "lumbar slouch" = "#D55E00",
    "shoulder slouch" = "#0072B2",
    "neck slouch" = "#CC79A7",
    "laying back" = "#F0E442",
    "unknown" = "white"
  )
  
  # Function to generate blend colors for combinations
  generate_blend <- function(postures, base_colors) {
    components <- unlist(strsplit(postures, "\\+"))
    if(length(components) == 1) return(base_colors[postures])
    
    # Convert hex to RGB
    rgb_vals <- col2rgb(base_colors[components])
    # Average the RGB values
    blend <- rowMeans(rgb_vals)
    # Convert back to hex
    rgb(blend[1], blend[2], blend[3], maxColorValue = 255)
  }
  
  # Create color mapping for all observed postures
  posture_colors <- sapply(all_postures, generate_blend, base_colors = base_colors)
  names(posture_colors) <- all_postures
  
  # Create label mapping (posture -> label with proportion)
  posture_labels <- proportions %>%
    mutate(label = paste0(classified_posture, " (", scales::percent(proportion, accuracy = 1), ")")) %>%
    select(classified_posture, label)
  
  # Create named vector for labels
  legend_labels <- posture_labels$label
  names(legend_labels) <- posture_labels$classified_posture

  # Create the plot
  plt <- time_ranges %>% 
    filter(participant == p, device_location == loc) %>%
    ggplot() +
      geom_rect(
        aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf, 
            fill = classified_posture),
        alpha = 0.4,
        color = NA,
        inherit.aes = FALSE
      ) +
      geom_line(
        aes(x = normalized_chip_time, y = normalized_angle_x),
        data = plot_data,
        color = "black"
      ) +
      scale_fill_manual(
        values = posture_colors,
        breaks = names(sort(posture_colors)),  # Sort for consistent legend
        labels = legend_labels,
        guide = guide_legend(
          title = "Posture Classification",
          nrow = 3,
          override.aes = list(alpha = 0.6)
        )
      ) +
      labs(
        x = "Normalized Time",
        y = "Normalized Angle",
        title = paste("Participant", p, "-", loc)
      ) +
      theme_minimal() +
      theme(
        legend.position = "bottom",
        legend.box = "vertical",
        panel.grid.major = element_line(color = "grey90"),
        panel.grid.minor = element_blank()
      )
  
  print(proportions %>% arrange(-proportion))
  
  plt  

}
```

#### Lower Back

```{r}
plot_classifications(free_norm_classified, time_ranges, 5, "lower_back")
```

#### Lower Back

```{r}
plot_classifications(free_norm_classified, time_ranges, 5, "mid_back")
```

#### Lower Back

```{r}
plot_classifications(free_norm_classified, time_ranges, 5, "neck")
```

### Ridge Plots

#### All Devices

```{r}
#| message: false
#| warning: false
free_norm %>%
  ggplot(aes(x = angle_x, y = participant, fill = device_location)) +
    geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
    scale_fill_viridis(
      option = "D", 
      discrete = TRUE,
      labels = c(
        "lower_back" = "Lower Back",
        "mid_back" = "Mid Back",
        "neck" = "Neck"
        )
      ) +
    labs(
      title = 'Distribution of Time Spent in Normalized Angle X by Participant',
      fill = 'Device Type',
      x = 'Angle X',
      y = 'Participant'
      ) +
    theme_ipsum() +
      theme(
        panel.spacing = unit(0.1, "lines"),
        strip.text.x = element_text(size = 8)
      ) +
    xlim(-50, 70)
```

#### Lower Back

```{r}
#| message: false
#| warning: false
free_norm %>%
  filter(device_location == 'lower_back') %>%
  ggplot(aes(x = angle_x, y = participant, fill = ..x..)) +
    geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
    scale_fill_viridis(option = "C") +
    labs(
      title = 'Distribution of Time Spent in Normalized Angle X by Participant', 
      subtitle = 'Lower Back',
      x = 'Angle X',
      y = 'Participant'
      ) +
    theme_ipsum() +
      theme(
        legend.position="none",
        panel.spacing = unit(0.1, "lines"),
        strip.text.x = element_text(size = 8)
      ) +
    xlim(-50, 70)
```

#### Mid Back

```{r}
#| message: false
#| warning: false
free_norm %>%
  filter(device_location == 'mid_back') %>%
  ggplot(aes(x = angle_x, y = participant, fill = ..x..)) +
    geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
    scale_fill_viridis(option = "C") +
    labs(
      title = 'Distribution of Time Spent in Normalized Angle X by Participant', 
      subtitle = 'Mid Back',
      x = 'Angle X',
      y = 'Participant'
      ) +
    theme_ipsum() +
      theme(
        legend.position="none",
        panel.spacing = unit(0.1, "lines"),
        strip.text.x = element_text(size = 8)
      ) +
    xlim(-50, 70)
```

#### Neck

```{r}
#| message: false
#| warning: false
free_norm %>%
  filter(device_location == 'neck') %>%
  ggplot(aes(x = angle_x, y = participant, fill = ..x..)) +
    geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
    scale_fill_viridis(option = "C") +
    labs(
      title = 'Distribution of Time Spent in Normalized Angle X by Participant', 
      subtitle = 'Neck',
      x = 'Angle X',
      y = 'Participant'
      ) +
    theme_ipsum() +
      theme(
        legend.position="none",
        panel.spacing = unit(0.1, "lines"),
        strip.text.x = element_text(size = 8)
      ) +
    xlim(-50, 70)
```



